---
title: "Estimating ecological change using the FIA database"
date: "`r Sys.Date()`"
author: "D. Perret"
output: 
  rmdformats::robobook:
    code_folding: hide
    self_contained: true
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      prompt = FALSE,
                      error = TRUE,
                      message = FALSE,
                      cache = FALSE,
                      cache.lazy = FALSE,
                      eval = TRUE,
                      eval.after = "fig.cap")

library(tidyverse)
library(rFIA)
library(ggplot2)
library(ggExtra)
library(rgdal)
library(sp)
library(raster)
select <- dplyr::select

# setting my preferred ggplot2 theme
theme_set(theme_bw())
theme_update(text = element_text(size=16, color = "black"),
             panel.grid.major = element_blank(),
             panel.grid.minor = element_blank(),
             strip.background = element_blank(),
             panel.border=element_rect(linewidth=1.5))

# colors to be used in all map figures
mapcolor <- "wheat3"
linecolor <- "gray40"
regioncolors <- c("Cascade Mixed Forest" = "#009E73",
                  "Northern Rocky Mountain Forest-Steppe" = "#56B4E9",
                  "Middle Rocky Mountain Steppe" = "#E69F00",
                  "Southern Rocky Mountain Steppe" = "#CC79A7",
                  "zIntermountain Semi-Desert" = "gray20",
                  "AZ-NM Mountains" = "#D55E00")

# sourcing estimation base code
source("growMort_rewrite_METRIC.R")


```


Loading data and modifying some fields:
```{r fia.data, echo=T}

# Another approach to loading and manipulating FIA data! Here we read in FIA data from a local directory, using functionality from the `rFIA` package (Stanke et al., 2019). This dataset was also downloaded with rFIA, using the rFIA::getFIA() function.

all.fia <- readFIA(dir = "data/fia_tables_all/",
                   common=T, states="OR")

#creating some fields in various tables

all.fia$PLOT <- all.fia$PLOT %>% 
  mutate(pltID = paste(UNITCD,STATECD,COUNTYCD,PLOT,sep="_"),
         PLT_CN = CN,
         ECOSUBCD = trimws(ECOSUBCD),
         state_key = "OR") %>% 
  group_by(pltID) %>% 
  mutate(most.recent = ifelse(MEASYEAR==max(MEASYEAR),
                              "yes","no")) %>% 
  ungroup()

## some summary dataframes and fields for convenience
abla.trees <- all.fia$TREE %>% 
  filter(SPCD == 19)

all.fia$PLOT <- all.fia$PLOT %>% 
  mutate(abla.pres = ifelse(PLT_CN %in% unique(abla.trees$PLT_CN),
                            1,0))

abla.plots <- all.fia$PLOT %>% 
  filter(abla.pres==1,
         most.recent=="yes",
         MEASYEAR>2009)

## creating some fields and updating all SPCDs to most-recently ID'd SPCD -- this is necessary because it's quite common for trees to change species ID, especially in smaller age classes.

all.fia$TREE <- all.fia$TREE %>% 
  left_join(all.fia$PLOT %>% 
              select(PLT_CN,most.recent,state_key),
            by="PLT_CN") %>% 
  mutate(DIA = DIA*2.54,
         PREVDIA = PREVDIA*2.54) # convert to metric 


# Loading some miscellaneous spatial data for plots and estimation

# WGS84
old.proj <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"

# Albers Equal Area; centered in western US
base.proj <- "+proj=aea +lat_0=23 +lon_0=-96 +lat_1=29.5 +lat_2=45.5 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"

# US state boundaries
states <- readOGR(dsn="data/spatial/",
                  layer = "state_boundaries", verbose=F) %>% 
  spTransform(.,
              CRSobj = CRS(base.proj))

# Subalpine fir range
range <- readOGR(dsn="data/spatial",
                 layer = "abielasi",
                 verbose = F,
                 p4s = old.proj) %>% 
  spTransform(.,
              CRSobj = CRS(base.proj))

# making a SpatialPointsDataFrame for subalpine fir plot locations (fuzzed)
abla.sp <- abla.plots %>% 
  SpatialPointsDataFrame(coords = .[,c("LON","LAT")],
                         data = .,
                         proj4string = CRS(old.proj)) %>% 
  spTransform(.,
              CRSobj = CRS(base.proj))

# Level 4 ecoregions (i.e., ecoregion subsections)
er4 <- readOGR(dsn="data/spatial",
               layer = "S_USA.EcomapSubsections",
               verbose=F) %>% 
  spTransform(.,
              CRSobj = CRS(base.proj))

er4.abla <- er4[abla.sp,]



```


Demonstrating how to pull out different change components:
```{r}

# very simple examples from tree table! 
## NOTE: the next database release will incorporate changes to the STATUSCD field; make sure that you're always reading the correct version of the user manual!

survivors <- all.fia$TREE %>% 
  filter(INVYR == 2019, # making sure we're only grabbing from the most recent remeasurements
         SPCD == 19,
         STATUSCD == 1,
         PREV_STATUS_CD == 1)

mortality <- all.fia$TREE %>% 
  filter(INVYR == 2019, # making sure we're only grabbing from the most recent remeasurements
         SPCD == 19,
         STATUSCD == 2,
         PREV_STATUS_CD == 1,
         AGENTCD != 80)

recruitment <- all.fia$TREE %>% 
  filter(INVYR == 2019, # making sure we're only grabbing from the most recent remeasurements
         SPCD == 19,
         STATUSCD == 1,
         is.na(PREV_STATUS_CD))

## We can pull out the same change components, with more nuance, using the TREE_GRM_COMPONENT table! To do that, you'll need to link the tables using the TRE_CN field.

```


Code snippets for how and where we assign components:

*Note:* For this analysis, we had to re-assign GRM components because we were doing a very large-scale analysis, and found inconsistencies between GRM treatments from different FIA units. This is why rFIA doesn't support GRM analyses for western states. Manually assigning GRM components in a consistent way across all states and units was a work-around to complete these analyses, and makes for a nice example of how this can be done:

```{r eval=F, echo=T}

## explanation of reCOMPONENT classification: 
## survivor/mortality1 = above thresh at both times
## survivor/mortality2 = crossed thresh
## survivor/mortality0 = stayed below thresh
## ingrowth1 = new tree, above thresh but still on microplot
## ingrowth2 = this is  "ongrowth" -- trees that would have been counted but was too small at t1. We generally *dont* want to include this! Presumed miscounted if grew directly onto subplot. Following FIA convention via Andy Gray.
## ingrowth0 = new tree, below thresh
## note: mortality includes harvest; agentcd==80

## ALSO NOTE : sizeThresh and DIA conditions differentiate between different components that get treated differently for time of selection and TPA factors

reCOMPONENT = case_when(
  #previous dead
  (STATUSCD==2 & PREV_STATUS_CD==2) | RECONCILECD==4 ~ "PREVDEAD",
  
  #full information
  STATUSCD==1 & PREV_STATUS_CD==1 & PREVDIA>=sizeThresh ~ "SURVIVOR1",
  STATUSCD==1 & PREV_STATUS_CD==1 & PREVDIA<sizeThresh & DIA>=sizeThresh ~ "SURVIVOR2",
  STATUSCD==1 & PREV_STATUS_CD==1 & DIA<sizeThresh ~ "SURVIVOR0",
  
  STATUSCD==2 & PREV_STATUS_CD==1 & PREVDIA>=sizeThresh ~ "MORTALITY1",
  STATUSCD==2 & PREV_STATUS_CD==1 & PREVDIA<sizeThresh & DIA>=sizeThresh ~ "MORTALITY2",
  STATUSCD==2 & PREV_STATUS_CD==1 & DIA<sizeThresh ~ "MORTALITY0",
  
  STATUSCD==1 & is.na(PREV_STATUS_CD) & RECONCILECD==0 & DIA>=sizeThresh & DIA<12.7 ~ "INGROWTH1",
  STATUSCD==1 & is.na(PREV_STATUS_CD) & RECONCILECD==0 & DIA>=sizeThresh & DIA>=12.7 ~ "INGROWTH2",
  STATUSCD==1 & is.na(PREV_STATUS_CD) & RECONCILECD==0 & DIA<sizeThresh ~ "INGROWTH0",
  STATUSCD==2 & is.na(PREV_STATUS_CD) & RECONCILECD==0 & DIA>=sizeThresh ~ "MORTALITY2",
  STATUSCD==2 & is.na(PREV_STATUS_CD) & RECONCILECD==0 & DIA<sizeThresh ~ "MORTALITY0",
  
  # reconciled trees
  RECONCILECD==1 & STATUSCD==1 & DIA>=sizeThresh & DIA<12.7 ~ "INGROWTH1",
  RECONCILECD==1 & STATUSCD==1 & DIA>=sizeThresh & DIA>=12.7 ~ "INGROWTH2",
  RECONCILECD==1 & STATUSCD==1 & DIA<sizeThresh ~ "INGROWTH0",
  RECONCILECD==1 & STATUSCD==2 & DIA>=sizeThresh ~ "MORTALITY2",
  RECONCILECD==1 & STATUSCD==2 & DIA<sizeThresh ~ "MORTALITY0",
  
  RECONCILECD==2 & STATUSCD==1 & DIA>=sizeThresh ~ "INGROWTH1",
  RECONCILECD==2 & STATUSCD==1 & DIA<sizeThresh ~ "INGROWTH0",
  RECONCILECD==2 & STATUSCD==2 & DIA>=sizeThresh ~ "MORTALITY2",
  RECONCILECD==2 & STATUSCD==2 & DIA<sizeThresh ~ "MORTALITY0",
  
  RECONCILECD==3 & STATUSCD==1 & DIA>=sizeThresh ~ "SURVIVOR1",
  RECONCILECD==3 & STATUSCD==1 & DIA<sizeThresh ~ "SURVIVOR0",
  RECONCILECD==3 & STATUSCD==2 & DIA>=sizeThresh ~ "MORTALITY1",
  RECONCILECD==3 & STATUSCD==2 & DIA<sizeThresh ~ "MORTALITY0",
  
  #unclassified
  TRUE ~ "UNCLASSIFIED"
)

```

Those GRM components are then used to ensure that the correct TPA expansion factor is used for each tree!

Estimation:
```{r}

### NOTE: this code is modified from Hunter Stanke's excellent `rFIA` package. See the included source code file, as well as `rFIA` documentation, and the very helpful `rFIA` tutorial.

evals <- c(41903,411903) #sampled plots in OR and WA used for GRM; see POP_EVAL table and EVALID variable for more info

dtpa.abla <- growMort_dlp.metric(db = all.fia,
                                 stateVar = "TPA", # we're counting trees, not calculating basal area or volume
                                 treeDomain = SPCD==19, # we're only interested in subalpine fir (SPCD==19)
                                 polys = er4.abla, # we're doing the estimation within Level IV ecoregions
                                 totals = T, 
                                 returnSpatial = T,
                                 sizeThresh = 2.54, # we're interested in all trees greater than 2.54 cm DBH
                                 evals = evals) %>% 
  filter(YEAR==2019)

```

MORTALITY
```{r}
ggplot() +
  geom_sf(data = states %>% 
            as(.,"sf"),
          fill=mapcolor,
          col=linecolor) +
  geom_sf(data = dtpa.abla %>% 
            filter(nPlots_AREA>9,
                   !is.na(CHNG_PERC)),
          col=NA,
          aes(fill = MORT_PERC)) +
  geom_sf(data = states %>% 
            as(.,"sf"),
          fill=NA,
          col=linecolor,
          alpha=0.1,
          lwd=0.3,
          lty=2) +
  theme(panel.background = element_rect(fill="skyblue1")) +
  scale_fill_steps(name = "",
                   #na.value="darkgray",
                   na.value=NA,
                   low = "white",
                   high = "firebrick3",
                   n.breaks = 6, limits = c(0,100))+
  lims(x = c(-2.4e6, -1.5e6),
       y = c(2.250e6,3.00e6)) +
  theme(legend.key.size = unit(1,"cm")) +
  labs(title = "Mortality (%)")

```

RECRUITMENT
```{r}
ggplot() +
  geom_sf(data = states %>% 
            as(.,"sf"),
          fill=mapcolor,
          col=linecolor) +
  geom_sf(data = dtpa.abla %>% 
            filter(nPlots_AREA>9,
                   !is.na(CHNG_PERC)),
          col=NA,
          aes(fill = RECR_PERC)) +
  geom_sf(data = states %>% 
            as(.,"sf"),
          fill=NA,
          col=linecolor,
          alpha=0.1,
          lwd=0.3,
          lty=2) +
  theme(panel.background = element_rect(fill="skyblue1")) +
  scale_fill_steps(name = "",
                   #na.value="darkgray",
                   na.value=NA,
                   low = "white",
                   high = "darkblue",
                   n.breaks = 6,
                   limits = c(0,100))+
  lims(x = c(-2.4e6, -1.5e6),
       y = c(2.250e6,3.00e6)) +
  theme(legend.key.size = unit(1,"cm")) +
  labs(title = "Recruitment (%)")

```

NET CHANGE
```{r}
ggplot() +
  geom_sf(data = states %>% 
            as(.,"sf"),
          fill=mapcolor,
          col=linecolor) +
  geom_sf(data = dtpa.abla %>% 
            filter(nPlots_AREA>9),
          col=NA,
          aes(fill = CHNG_PERC)) +
  geom_sf(data = states %>% 
            as(.,"sf"),
          fill=NA,
          col=linecolor,
          alpha=0.1,
          lwd=0.3,
          lty=2) +
  theme(panel.background = element_rect(fill="skyblue1")) +
  scale_fill_steps2(name = "ABLA % change \n(abundance)",
                    na.value=NA,
                    low = "firebrick3",
                    mid="white",
                    midpoint=0,
                    high = "darkblue",
                    limits = c(-100,100),
                    n.breaks = 9)+
  lims(x = c(-2.4e6, -1.5e6),
       y = c(2.250e6,3.00e6)) +
  theme(legend.key.size = unit(1,"cm")) +
  labs(title = "Abundance change (%)")

```
